---
title: "HIV Analysis"
author: "STA198 Team: Natalie, Ella, Jessica, Ali"
date: "2026-04-10"
output: pdf_document
---

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
library(janitor)

deaths <- read_csv("Data/no_of_deaths_by_country_clean.csv")
hiv <- read_csv("Data/no_of_people_living_with_hiv_by_country_clean.csv")
art <- read_csv("Data/art_coverage_by_country_clean.csv")
```

```{r}
deaths_clean <- deaths |>
  clean_names() |>
  group_by(country) |>
  summarise(
    deaths = first(count_median),
    .groups = "drop"
  )

```

```{r}
hiv_clean <- hiv |>
  clean_names() |>
  group_by(country) |>
  summarise(
    hiv_population = first(count_median),
    .groups = "drop"
  )
```

```{r}
art_clean <- art |>
  clean_names() |>
  select(
    country,
    art_coverage = estimated_art_coverage_among_people_living_with_hiv_percent_median
  )

```

```{r}
final_df <- deaths_clean |>
  inner_join(hiv_clean, by = "country") |>
  inner_join(art_clean, by = "country")
```

```{r}
write_csv(final_df, "Data/final_df.csv")
```

```{r}
glimpse(final_df)
```

## Description
I created a merged, final dataset to use for analysis with cleaned and standardized columns (contains 170 countries and 4 variables). Each row represents a single country. The dataset was constructed by merging three country-level datasets sourced from WHO-based HIV/AIDS indicators available on Kaggle. All variables are continuous numerical estimates, and represent median values reported in the original datasets. The variables included in the final dataset are:
- HIV/AIDS deaths (deaths)
- Number of people living with HIV (hiv_population)
- ART coverage among people living with HIV (art_coverage)




```{r}
colSums(is.na(final_df))

```
```{r}
colMeans(is.na(final_df))
```

## Description
The dataset contains missing values across all three main variables: HIV/AIDS deaths (37 missing, ~21.8%), number of people living with HIV (32 missing, ~18.8%), and ART coverage (34 missing, 20%); no missing data for countries. This is most likely due to incomplete reporting in certain countries within the original WHO-based dataset (real-world data is often incomplete). Missing data was not cleaned or removed, just be wary of it and decide what to do with it in the next stages (if anything).





```{r}
final_df |> 
  arrange(desc(deaths)) |> 
  slice(1:10)
```
## Description
Shows top 10 countries with the most HIV/AIDS deaths.

```{r}
final_df |> 
  filter(!is.na(deaths)) |>
  arrange(deaths) |>
  slice(1:10)
```

## Description
Shows top 10 countries with the least HIV/AIDS deaths.




```{r}
ggplot(final_df, aes(x = deaths)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of HIV/AIDS Deaths", x = "Deaths", y = "Count of Countries")
```
## Description
Shows the distribution of deaths across countries.




```{r}

ggplot(final_df, aes(x = art_coverage)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of ART Coverage", x = "ART Coverage (%)", y = "Count of Countries")

```
## Description
Shows the distribution of treatment access (ART coverage %) across countries.




```{r}
ggplot(final_df, aes(x = art_coverage, y = deaths)) +
  geom_point() +
  scale_y_log10()
```
## Description
Shows the relationship between ART coverage (%) and HIV/AIDS deaths across countries, where each point represents a country. Note that I used log scaling for the y-axis to account for scaling issues and compression of the graph.

