---
title: "HIV Analysis"
author: "STA198 Team"
date: "2026-04-10"
output: pdf_document
---

```{r}
rm(list = ls())
```

```{r}
library(tidyverse)
library(janitor)

deaths <- read_csv("Data/no_of_deaths_by_country_clean.csv")
hiv <- read_csv("Data/no_of_people_living_with_hiv_by_country_clean.csv")
art <- read_csv("Data/art_coverage_by_country_clean.csv")
```

```{r}
deaths_clean <- deaths |>
  clean_names() |>
  group_by(country) |>
  summarise(
    deaths = first(count_median),
    .groups = "drop"
  )

```

```{r}
hiv_clean <- hiv |>
  clean_names() |>
  group_by(country) |>
  summarise(
    hiv_population = first(count_median),
    .groups = "drop"
  )
```

```{r}
art_clean <- art |>
  clean_names() |>
  select(
    country,
    art_coverage = estimated_art_coverage_among_people_living_with_hiv_percent_median
  )

```

```{r}
final_df <- deaths_clean |>
  inner_join(hiv_clean, by = "country") |>
  inner_join(art_clean, by = "country")
```

```{r}
summary(final_df)

```

## Description
I created a merged, final dataset to use for analysis with cleaned and standardized columns (contains 170 countries and 4 variables). Each row represents a single country. The dataset was constructed by merging three country-level datasets sourced from WHO-based HIV/AIDS indicators available on Kaggle. All variables are continuous numerical estimates, and represent median values reported in the original datasets. The variables included in the final dataset are:
- HIV/AIDS deaths (deaths)
- Number of people living with HIV (hiv_population)
- ART coverage among people living with HIV (art_coverage)












